---
title: "Covid Data ExploRation"
output:
  word_document: default
  html_document: default
  pdf_document: default
date: "2023-09-07"
editor_options:
  markdown:
    wrap: 72
---

## Install Packages and set up environment

#### When knitting a markdown document install packages in console and libraries in the markdown

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

install.packages("tidyverse") 

install.packages("ggplot2") 

install.packages("googledrive") 

drive_auth([email="youremail@gmail.com"])

```{hide}
drive_auth([email="k.canivet.usa@gmail.com]")
```
To make executable code in a "Chunk" add "\`\`\`{r}"
```{r librarys, message=FALSE}
library(tidyverse)
library(ggplot2)
library(googledrive)
```

## Project Goals

#### The goal of this project is to replicate a workflow I conducted in SQL based off a tutorial by "Alex the Analyst"

[https://youtu.be/qfyynHBFOsM?si=Nzr2eiFWr_3gs7w6)](https://youtu.be/qfyynHBFOsM?si=Nzr2eiFWr_3gs7w6)){.uri}

### Getting the data

#### I had the data stored in a google drive folder,so I am accessing it from there

Authorize R Studio to communicate with Google Drive

Navigate to the Google Drive folder that contains the data and list the
files

```{r file_list, message=FALSE, warning=FALSE}
drive_ls("~/Portfolio/Covid19")
```

Get source datafile from Google Drive and open in R Studio

```{r Pull_Data_File, eval=FALSE, include=FALSE}
drive_download("owid-covid-data.csv")

```

Read the csv into a dataframe

```{r loadcsv}
covid_data_full <- read_csv("owid-covid-data.csv")
```

#### Inspect the data frame

```{r head}
head(covid_data_full)
```

```{r columnlist}
colnames(covid_data_full)
```

#### Create 2 separate dataframes, one for cases and deaths, one for

vaccination data

```{r filesplit}
OWID_Deaths <- subset(covid_data_full, select = c(iso_code,continent, location, date,population,total_cases,new_cases,total_deaths,new_deaths))

OWID_Vax <- subset(covid_data_full, select = c(iso_code, continent,location,date,total_vaccinations,new_vaccinations,people_vaccinated,people_fully_vaccinated, total_boosters))

```

Verify columns in new dataframes

```{r dataframes}
colnames(OWID_Deaths)
```

Filter the OWID_Deaths dataframe to only show a select list of countries

```{r select_countries}
OWID_Deaths_filter_location <- OWID_Deaths %>% 
  filter(location %in% c("Canada", "United States", "Italy","Brazil"))

show(OWID_Deaths_filter_location)
```

An advantage of using R is adding charts for a quick look. Adding a line
chart to compare selected countries with total deaths over time

```{r country_graph, message=FALSE}
ggplot(data = OWID_Deaths_filter_location)+
  geom_line(mapping=aes(x=date,y=total_deaths,color=location))
```

```{r head_vax}
head(OWID_Vax)
```

Add a new calculated column to the OWID_Deaths dataframe to get a
percentage. Then check the calculation by filtering when the United
States had values greater than 1% of the population

```{r OWID_Deaths}
OWID_Deaths <- OWID_Deaths %>%
  mutate(percent_pop=(total_cases/population)*100)

print(OWID_Deaths) %>% 
  filter(!is.na(total_cases), location=="United States", percent_pop>1) %>% 
  slice(1:25)
```

show the maximum percentage from the OWID_Deaths dataframe for each
country, ranked from highest to lowest

null results can't mix with numerical values for summarize. The 2nd filter step is necessary before summarize
```{r Deaths_rank}
OWID_Deaths_Max <- OWID_Deaths %>% 
  select(continent,location,total_deaths) %>% 
  filter(!is.na(continent)) %>% 
  group_by(location,continent) %>% 
  filter(!is.na(total_deaths)) %>% 
  summarize(max_deaths=max(total_deaths)) %>% 
  arrange(desc(max_deaths))  
  
print(OWID_Deaths_Max)
```

Rank countries by percent of population with Covid19

```{r perc_rank}
OWID_Deaths_MaxP <- OWID_Deaths %>% 
  select(continent,location,percent_pop) %>% 
  filter(!is.na(percent_pop)) %>% 
  group_by(location,continent) %>% 
  summarize(max_prec=max(percent_pop)) %>% 
  arrange(desc(max_prec)) 
 
print(OWID_Deaths_MaxP)
```

Join the two OWID dataframes

```{r join}
joined_OWID <- left_join(OWID_Deaths,OWID_Vax, by = c("location","date" ))

colnames(joined_OWID)
```

```{r}
head(joined_OWID)
```

create a rolling count

```{r rolling_count}
joined_OWID_RC <- joined_OWID %>%
  arrange(location, date) %>% 
  filter(!is.na(continent.x)) %>% 
  group_by(location,continent.x) %>% 
  filter(!is.na(new_vaccinations)) %>% 
  mutate(Rolling_Sum_Vax = cumsum(new_vaccinations)) %>% 
  filter(location %in% c("Canada", "United States","Brazil"))

print(joined_OWID_RC[, c("continent.x","location","date", "Rolling_Sum_Vax")])
```

Take the previous dataframe and rank the countries by the total of new
vaccinations

```{r total_summary}
joined_OWID_RC_totals <- joined_OWID %>%
  arrange(location, date) %>% 
  filter(!is.na(continent.x)) %>% 
  group_by(location,continent.x) %>% 
  filter(!is.na(new_vaccinations)) %>% 
  mutate(Rolling_Sum_Vax = cumsum(new_vaccinations)) %>% 
  #filter(location %in% c("Canada", "United States","Brazil"))
  summarize(rolling_total=max(Rolling_Sum_Vax)) %>% 
  arrange(desc(rolling_total)) 
  
print(joined_OWID_RC_totals)
```

Trim the dataframe to only show the top 10

```{r top10}
joined_OWID_RC_top10 <- joined_OWID_RC_totals %>%
    arrange(desc(rolling_total)) %>% 
     head(10)

print(joined_OWID_RC_top10)
```

### Show some charts of the data

```{r column_chart}
ggplot(data=joined_OWID_RC_top10) +
         geom_col(mapping=aes(x=reorder(location,rolling_total),y=rolling_total,fill=location,width = 0.9))+
          coord_flip()+
          labs(title = "Rolling Total Vaccinations",subtitle="Top 10",y="Total Vaccinations (Billions)",x="Country")+
          scale_y_continuous(labels = scales::comma_format(scale = 1e-9))
```
